---
title: "Gov 50 Final Project"
author: Austen Wyche
description: Final Project for GOV50
output:
  distill::distill_article:
---



## Project thoughts

I am interested in exploring data related to the relationship between gun ownership rates in the United States and the frequency of mass shootings. Does increased gun ownership have a positive relationship with the frequency of mass shootings, or is there no relationship between the two?

## Milestone 2

Do gun ownership rates correlate with the number of mass shootings in the United States? In this study, I plan to analyze the relationship between self-proclaimed Gun Ownership in Gallup polling and the occurance of mass shootings (defined by a shooting where at least three deaths occur). I hypothesize there is likely no relationship between gun ownership rates and mass shootings. I expect that the rate of gun ownership in the United States has remained stable, while the rate of mass shootings has likely increased over the past couple of decades. My unit of analysis is the United States over the period 1980-2017. The explanatory variable of interest is the gun ownership rate, measured by public policy polling. The outcome variable is the number of mass shootings in the United States. The outcome variable of interest is measured by Mother Jones, a nonprofit organization which compiled records of mass shootings in the United States since the year 2000. A pattern that displays little to no correlation between gun ownership rates and mass shootings in the United States would confirm my hypothesis. 

## Milestone 3

```{r}
library(tidyverse)
library(ggplot2)
library(gov50data)
library(infer)
library(dbplyr)
gun_death_data <- read.csv("massshootings.csv")
```

```{r}
library(dplyr)
library(ggplot2)

gun_deaths_by_year <- gun_death_data|>
  group_by(Year)|>
  summarize(n = n())

gun_plot <- gun_deaths_by_year %>%
  ggplot(mapping = aes(x = Year, y = n)) +
  geom_bar(stat = "identity")+
  labs(title = "Number of Mass Shootings from 1980 to 2017", x = "Year", y = "Number of Mass Shootings")
gun_plot

```


```{r}
library(tidyverse)
library(ggplot2)
library(gov50data)
library(infer)
library(dbplyr)

gun_ownership_data = read_csv("gunownership.csv")
gun_ownership_data

gun_ownership_data|>
  select(Year, OwnGuns)|>
  ggplot(mapping = aes(x = Year, y = OwnGuns))+
  geom_col()+
  labs(title = "Gallup Gun Ownership Rate in the United States from 2000-2018", x = "Years", y = "Gun Ownership Rate(%)")
```


## Milestone 4

```{r}
merged_data <- gun_death_data|>
  inner_join(gun_ownership_data, by = "Year")|>
  select(OwnGuns, Total.victims, Fatalities, Year)
```

```{r}
gun_ownership_data
fit <- lm(data= gun_ownership_data, OwnGuns~Year)|>
  broom::tidy()|>
  select(term, est=estimate, se=std.error, pval = p.value)

knitr::kable(fit, digits = 3, col.names = c("Variable", "Coefficient Estimate", "Standard Error", "P Value"))

ggplot(data= gun_ownership_data, aes(x= Year, y = OwnGuns))+
  geom_col()+
  geom_smooth(method="lm")
```

My linear regression can be interpreted as meaning for every one unit increase (year), there is an increase in the gun ownership rate by an average .26 percentage points. As the p value is under .01, it can be seen as statistically significant. Lastly, the intercept should not be interpreted causally as it begins at Year Zero, when guns were not yet invented. 

